Trawl Summary

### 2020-05-11 text/trawl_summary.Rmd
updated trawl_summary.Rmd file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(vegan)
library(betapart)
library(lubridate)
library(viridis)

@@

More data will be necessary to fully complete analyses, but this shows the emerging environmental picture.

@@

The education department has been conducting trawl surveys pretty consistently since 2007. These sampling dates are pretty evenly distributed throughout the year

@@

There are a lot of sampling trawls over the past decade (n = 359) and they are spread pretty consistently over the year, but with some periods of heavy sampling:

@@

Over the full dataset we can estimate the species accumulation by # of trawls

@@

And by the trawls within each year to get at the differences in sampling effort (assuming each trawl is the same). This suggests there are certain years (or trawl locations):

```{r, include = TRUE, echo = FALSE, warning=FALSE}
## SAC for each year
TB_trawl_commondate <- TB_trawl_data %>%
  select(year, Date, Common_name) %>%
  group_by(Date, Common_name) %>%
  distinct() %>% 
  mutate(pres = 1) %>%
  pivot_wider(names_from = Common_name, values_from = pres, values_fill = list(pres = 0)) %>%
  ungroup()

@@

Squinting this suggests that, generally, there might be increasing # of species more broadly.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(vegan)
library(betapart)
library(lubridate)
library(viridis)
library(gridExtra)

@@

There are a number of reasons for data holes and more complete data will be necessary for any final analyses, but this shows the emerging environmental picture.

@@

The education department has been conducting trawl surveys pretty consistently since 2007.

@@

There are a lot of sampling trawls over the past decade (n = 359) and they are spread over the year, with some periods of heavy sampling:

@@

Over the full dataset we can estimate the species accumulation by # of trawls from the `specaccum()` function in the <a href = "https://cran.r-project.org/web/packages/vegan/index.html" target = "_blank">vegan package</a>

@@

The trawls within each year give some insight into the differences species-effort through time (assuming each trawl is the same). This suggests there are certain years (or trawl locations) that have higher richness:

```{r SAC curves, include = TRUE, echo = FALSE, warning=FALSE, }
## SAC for each year
TB_trawl_commondate <- TB_trawl_data %>%
  select(year, Date, Common_name) %>%
  group_by(Date, Common_name) %>%
  distinct() %>% 
  mutate(pres = 1) %>%
  pivot_wider(names_from = Common_name, values_from = pres, values_fill = list(pres = 0)) %>%
  ungroup()

@@

Squinting this suggests that, generally, there might be increasing # of species based on the species-effort curves. The 'brighter' colors appear to be higher. I think we still need to dig in further to see if this is due to changing taxonomic resolution across the sampling period or an actual biological signal.

## Beta diversity

We can start to break this down further, and the inference that species richness might be increasing over time has some support (again, not sure yet if this is real or methodological).

```{r beta partition, include = TRUE, echo = FALSE, warning = FALSE, message= FALSE, fig.height= 8}

#create time-accumulation df from trawl data
TB_trawl_data %>%
  filter(!duplicated(Common_name)) %>%
  mutate(richness = 1:n(), 
         Date = as.Date(Date),
         richness = sort(richness))-> TB_time_accum

TB_annual_accum <- TB_trawl_data %>%
  mutate(pres = 1) %>% group_by(year) %>%
  filter(!duplicated(Common_name)) %>%
  summarise(richness = sum(pres)) %>%
  mutate(Date = as.Date(as.character(year), format = "%Y", origin = "2000-01-01"))

# partition beta diversity into nestedness and turnover #
# uses `betapart` package #
TB_time_df <- TB_time_accum %>%
  select(Date, year, richness) %>%
  mutate(data_type = 'continuous') %>%
  bind_rows(TB_annual_accum %>% mutate(data_type = "annual"))

#create moving jaccard calc by year-month
# create a dataframe for time series of distance compared to first sampling 
unique_combinations <- expand.grid(date_A = unique(levels(as.factor(TB_trawl_commonsite$date_id))), 
                                   Date = unique(levels(as.factor(TB_trawl_commonsite$date_id)))) %>%
  filter(date_A == as.character(min(TB_trawl_commonsite$date_id)))

TB_date_jaccard = vegdist((TB_trawl_commonsite %>% select(-year, -month) %>% 
                             column_to_rownames("date_id")), method = "jaccard") %>%
  as.matrix %>% as.data.frame %>% rownames_to_column("date_A") %>% 
  pivot_longer(cols = 2:dim(.)[2], names_to = "Date", values_to = "jaccard") %>%
  inner_join(unique_combinations) %>% filter(date_A != Date) %>% 
  mutate(Date = gsub('^([0-9]{4}-)([0-9]{1})$', '\\10\\2', Date),
         Date = as.Date(paste0(Date,"-01"), format = "%Y-%m-%d"))

# partitioning beta diversity metrics from trawl dataset
# create named list of each sampling date
named_group_split <- function(.tbl, ...) {
  grouped <- group_by(.tbl, ...)
  names <- rlang::eval_bare(rlang::expr(paste(!!!group_keys(grouped), sep = " / ")))

  grouped %>%
    group_split() %>%
    rlang::set_names(names)
}

date_lists = TB_trawl_commonsite %>%
  named_group_split(date_id) %>%
  map(., ~.x %>% select(-year, -month,-date_id))

beta_lists <- lapply(date_lists, function(x,r){ betapart::beta.temp(x = x, y = r, index.family = 'jaccard')}, x = date_lists[[1]]) 
beta_dates <- gsub('^([0-9]{4}-)([0-9]{1})$', '\\10\\2', names(beta_lists))
beta_df <- bind_rows(beta_lists) %>% mutate(Date = as.Date(paste0(beta_dates,"-01"), "%Y-%m-%d")) %>% #select(Date, everything()) %>%
  pivot_longer(-Date, names_to = "partition", values_to = "value")

rich_time <- 
  ggplot(TB_time_df, aes(x = Date, y = richness, group = data_type)) +
  geom_point(aes(colour = data_type),size = 1.5) +
  geom_line(aes(colour = data_type),size = 1.2) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  scale_y_continuous(name = "Species Richness")+
  scale_colour_manual(name = "", values = viridis(4)[c(2,3)]) +
  theme_minimal() +
  theme(legend.position = c(0, 1), legend.justification = c(0, 1), legend.background = element_rect(fill = "transparent", colour = NA),
        axis.title.x = element_blank(), axis.text.x = element_blank(),
        legend.spacing.y = unit(-0.1, 'cm'))

 jaccard_time <-
  ggplot(beta_df, aes(x = Date, y = value, group = partition)) + 
   geom_point(aes(colour = partition), size = 1.5) +
  geom_line(aes(colour = partition), size = 1.2) +
  scale_y_continuous(name = "Jaccard dissimilarity", limits = c(0,1.1), expand = c(0.01,0.01))+
  scale_colour_manual(name = "", values = viridis(7)[c(2,4,7)], labels = c("jaccard", "nestedness","turnover")) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
   theme_minimal() +
  theme(legend.background = element_rect(colour = NA, fill = 'transparent'),
        legend.position = c(0,1), legend.justification = c(0,1), axis.title.x = element_blank(),
        legend.spacing.y = unit(-0.2, 'cm'))

grid.arrange(rich_time, jaccard_time, ncol = 1)

The bottom panel can use a bit more explanation. This panel shows jaccard distance of each individual trawl from the initial trawl based on presence/absence data. Understanding the highlights and limitations of the analysis can give us a better picture of what is going on. First, a community in flux responding to directional change, we would expect the distance to increase over time. There is maybe slight evidence of this, but a lot of noise around that signal. Further, the relative effects of species turnover is quite variable over time and seems to show a larger seasonal signal relative to any trend over the time series.

Taking this further,
